Higher-Order unification for free!

### 2024-07-24 paper/paper_for_taps/source/main.tex
add paper_for_taps folder (to build the zip)

\usepackage{myTools}
\usepackage{macros}
\usepackage{enumitem}
\usepackage{intcalc}
\usepackage{nameref}

% TODO: set this fields
%\setcopyright{cc}
%\setcctype{by}
%\copyrightyear{2024}
%\acmYear{XXXX 2024}
%\acmBooktitle{YYY}
%\acmDOI{ZZZZZZZZZZZZ}

@@

In order to reason about unification we provide a description of the
\Fo{} and \Ho languages where unification variables
are first-class terms, i.e. they have a concrete syntax as
shown in \cref{code:common-terms}.
Unification variables
in \Fo{} (\elpiIn{o-uva} term constructor) have no explicit scope:
the ``arguments'' of a higher-order variable are given via the \elpiIn{o-app}
constructor. For example the term <<\coqIn{P x}>> is represented as
<<\elpiIn{o-app [o-uva N, x]}>>, where \elpiIn{N} is the memory address
of \elpiIn{P} and \elpiIn{x} is a bound variable.
In \Ho{} the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
since unification variables are higher-order and come equipped with an
explicit scope.
%
% We keep these languages
%minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
%in the example in Section~\ref{sec:intro} together with the type notation of
%terms carried by the \elpiIn{lam} constructor.
%
{
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}
\begin{figure}[b] %[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
   \inputrawelpicode{code/fo_tm}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
   \inputrawelpicode{code/ho_tm}
  \end{minipage}
  \end{tabular}\vspace{4pt}
  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}
}
%\todo{use faddr for Fo}
%
%faddr and addr are memory addresses, the details are given in bla,
%here they are just unique identifiers for unif variables and their types
%do now allow to mistake one for the other.
%\todo{cleanup}

@@

% An example of $t$ in \notllambda{} is $F \appsep a$ for a constant $a$.
% Note however tha
% an oracle could provide an assignment $\rho = \{ F \mapsto \lambda x.x\}$
% that makes the resulting term fall back in \llambda.
% \todo{said before}

\newcommand{\subterm}[1]{\ensuremath{\mathcal{P}(#1)}}
We write $\subterm{t}$ for the set of sub-terms of $t$, and
%\begin{definition}[Subterms \subterm{t}] The set of sub terms of $t$ is the
%  largest set $\subterm{t}$ that can be obtained by the following rules.
%$$
%\begin{array}{l}
%t \in \subterm{t}\\
%t = f t_1\ldots t_n \Rightarrow \subterm{t_i} \subseteq \subterm{t} \land f \in \subterm{t}\\
%t = \lambda x.t' \Rightarrow \subterm{t'} \subseteq \subterm{t}\\
%\end{array}
%$$
%\end{definition}
%\noindent
we write $\subterm{X} = \bigcup_{t\in X} \subterm{t}$ when $X$ is a set of terms.

@@

% \noindent
% \todo{remove pruning optimization from code above}

@@

\begin{lemma}[$\wellb$-partial-enforcement]\label{lem:w-enforcement-maybebeta}
  $\forall s, \C{s}{t}{m}{l}$, if $\:\wellb_\beta(s)$  then $\:\wellb(t)$.
\end{lemma}
% \begin{proof}[Proof sketch]
%   problematic terms are mapped to uva by comp,
%   the problematic o-app node is gone.
%   if F x = t then  F = lam x.t[]
%   \todo{finish/maybe W enforcement is not needed here}
% \end{proof}
\noindent
In other words the compiler eliminates all subterms that are in \maybebeta.

@@

\begin{theorem}[Fidelity in $\wellb_{\beta \eta}$]
  % F x = z, F x y = z y
  Given a list of unification problems \foUnifPb such that
  $\wellb_{\beta \eta}(\foUnifPb)$, if the 
  memory map is bijective then
  the introduction 
  of \linketa guarantees \fullref{prop:fidelity}.
  %\footnote{The bijective property of \mapStore is proved in \cref{sec:invariant1}}
  \label{lemma:fidelity-maybeeta}
\end{theorem}

\begin{proof}[Proof sketch] %\todo{to reformulate}
  % For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  % succeeds. A $\hstep_i$ is made by a unification
  % $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
  % progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
  We want to prove that \fstep succeeds iff \hstep
  succeeds, where \hstep is made by a unification step $u$ and a link
  progression $p$.
  Without loss of generality we consider a
  \Fo{} unification problem of the form $s_1 \Uo s_2$
  where $s_1 \in \maybeeta$ and is compiled to a variable $X$ with the accessory link
  $\linketaM{\Gamma}{X}{\lambda x.t_1}$. 
  The unification $u$ always succeeds, since $X$ is a fresh variable, we 
  only have to consider the link progression $p$.
  Two cases should be analyzed:
  \begin{itemize}
    \item $s_2 \in \maybeeta$, in this case, the unification problem becomes $X \Ue Y$ 
          with the extra link $\linketaM{}{Y}{\lambda x.t_2}$. After the unification of 
          $X$ and $Y$, \progressetadedup is fired and, once the lambdas are crossed,
          $t_1$ and $t_2$ are unified. 
          This unification succeeds iff $s_1 \Uo s_2$ succeeds by \cref{th:fidelity-beta}.

@@

% \todo{explain why decompile does so and which eta?}

\begin{comment}
\paragraph{Example of \progressetadedup}

A very basic example of \linketa deduplication, is given below:
% make test ONLY=7002 TEX=tex
\printAlll
  {{{\lambda x.(X\appsep x),\lambda x.(Y\appsep x)}}}
  {{{A,C}}}
  {{{X,B,1},
    {Y,D,1}}}
  {{{\eta,,A,\lambda x.B_{x}},
    {\eta,,C,\lambda x.D_{x}}}}

\noindent
The result of $A \Ue{} C$ is that the two \linketa share the same \lhs.
By unifying the two \rhs we get
$\sigma = \{ A~ \mapsto C, B \mapsto D ~\}$.
In turn, given the map \mapStore, this second assignment is decompiled to
$\rho = \{~ X \mapsto Y ~\}$ as expected.
\end{comment}

%\section{Enforcing inv.: \NoCaseChange{\nameref{inv:uvaarity}}}
\section{Making \mapStore a bijection}
\label{sec:invariant1}

We want to allow for partial application of functions.
As a consequence
the same unification variable $X$ can be used multiple times
with different arities.
When it is the case the
memory map \mapStore generated by the compiler is not
a bijection. For example:
\begin{comment}
%
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}
\end{comment}
% make test ONLY=7007 TEX=tex
\printAlll
  {{{\lambda x.(X\appsep x),f},
    {X,\lambda x.a}}}
  {{{A,f},
    {C,\lambda x.a}}}
  {{{X,C,0},
    {X,B,1}}}
  {{{\eta,,A,\lambda x.B_{x}}}}
% \todo{simplify example since no prog right}
in the unification problems \foUnifPb above, $X$ is used with arity $1$ in
$\foUnifPb_1$ and with arity $0$ in $\foUnifPb_2$. 
In order to preserve \fullref{inv:uvaarity} the compiler
did generate two entries in \mapStore for $X$, namely $B$ and $C$. 
However, there is no connection between these two variables
and any incompatible assignements to them would not be detected,
in turn breaking \cref{prop:fidelity}. 
% This incompatibility will only be detected during the
% decompilation phase.
To address this issue, we post-process \mapStore
to ensure the following property:

@@

% \todo{this proof is still unreadable}
% \todo{Not sure I fullfy fixed the proof, but the old oen was not ok}

% Note that shortening the chain of links allows unification to fail if
% $X_1$ and $X_2$ are unified with different terms.
% \todo{what are these X1 and X2?}

@@

\begin{comment}
\paragraph{Example of \progBetaLL}
Consider the \linkbeta below:
%
% make test ONLY=7004 TEX=tex
\printAlll
  {{{X,\lambda x.x},
    {\lambda x.(Y\appsep (X\appsep x)),f}}}
  {{{A,\lambda x.x},
    {B,f}}}
  {{{Y,D,0},
    {X,A,0}}}
  {{{\eta,,A,\lambda x.E_{x}},
    {\eta,,B,\lambda x.C_{x}},
    {\llambda,x,C_{x},(D\appsep E_{x})}}}

\noindent
Initially the \linkbeta \rhs is a variable $D$ applied to the $E_x$.
The first unification problem results in $\sigma =
\{A \mapsto \lambda x.x\}$. In turn this instantiation
triggers $\linkStore_1$ by \progressetaleft 
and $E_x$ is assigned to $x$.
Under this substitution the \linkbeta becomes
\linkbetaM{x}{C_x}{(D \appsep x)}, and by \progBetaLL
it is replaced with the link:
\linketaM{}{E}{\lambda x.D_x}, while $C_x$ is unified with $D_x$. The second unification
problem assigns $f$ to $B$, that in turn activates
the second \linketa ($f$ is assigned to $C$), and then all the remaining links
are solved. The final \Ho substitution is $\sigma = \{~A _{} \mapsto \lambda x.x ~;~
B _{} \mapsto f ~;~ C _{x} \mapsto (f\appsep x) ~;~
D _{} \mapsto f ~;~ E _{x} \mapsto x ~;~
F _{x} \mapsto C_{x} ~\}$ and is decompiled into $\rho = \{~X \mapsto \lambda x.x ~;~
Y \mapsto f~\}$.

\paragraph{Example of \progBetaRH}
We give a second look at the opening example in \cref{sec:beta}.
% \todo{this is a little more complex that the opening problem}
The problem is compiled into:
%
% make test ONLY=7003 TEX=tex
\printAlll
  {{{X,\lambda x.a},
    {(X\appsep a),a}}}
  {{{A,\lambda x.a},
    {B,a}}}
  {{{X,A,0}}}
  {{{\llambda,,B,(A\appsep a)}}}

% \todo{Rinominare i print dei problemi}

The first unification problem is solved with the substitution
$\sigma = \{~A \mapsto \lambda x.a~\}$. The \linkbeta becomes
\linkbetaM{}{B}{((\lambda x.a) \appsep a)} whose \rhs can be $\beta$-reduced to
$a$. $a$ is in \llambda and is unified with $B$. The resolution of the second
unification problem
confirm that $B$ equal $a$. The final substitution $\sigma = \{~A \mapsto \lambda x.a ~;~ B \mapsto a ~\}$ is decompiled into $\rho = \{~ X \mapsto \lambda x.a ~;~ Y \mapsto
a ~\}$.
\end{comment} 
\subsection{\texorpdfstring{Relaxing \fullref{def:progBetaFail}}{Relaxing progress-llam-fail}}
\label{sec:heuristics}
\newcommand{\progBetaNoLLWait}{\emph{progress-beta-\notllambda}}

>>>>

\usepackage{myTools}
\usepackage{macros}
\usepackage{enumitem}

@@

In order to reason about unification we provide a description of the
\Fo{} and \Ho languages where unification variables
are first-class terms, i.e. they have a concrete syntax as
shown in \cref{code:common-terms}.
Unification variables
in \Fo{} (\elpiIn{o-uva} term constructor) have no explicit scope:
the ``arguments'' of a higher-order variable are given via the \elpiIn{o-app}
constructor. For example the term <<\coqIn{P x}>> is represented as
<<\elpiIn{o-app [o-uva N, x]}>>, where \elpiIn{N} is the memory address
of \elpiIn{P} and \elpiIn{x} is a bound variable.
In \Ho{} the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
since unification variables are higher-order and come equipped with an
explicit scope.
%
% We keep these languages
%minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
%in the example in Section~\ref{sec:intro} together with the type notation of
%terms carried by the \elpiIn{lam} constructor.
%
{
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}
\begin{figure}[b] %[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
   \inputrawelpicode{code/fo_tm}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
   \inputrawelpicode{code/ho_tm}
  \end{minipage}
  \end{tabular}\vspace{4pt}
  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}
}

@@

\newcommand{\subterm}[1]{\ensuremath{\mathcal{P}(#1)}}
We write $\subterm{t}$ for the set of sub-terms of $t$, and
we write $\subterm{X} = \bigcup_{t\in X} \subterm{t}$ when $X$ is a set of terms.

@@

% \noindent

@@

\begin{lemma}[$\wellb$-partial-enforcement]\label{lem:w-enforcement-maybebeta}
  $\forall s, \C{s}{t}{m}{l}$, if $\:\wellb_\beta(s)$  then $\:\wellb(t)$.
\end{lemma}
\noindent
In other words the compiler eliminates all subterms that are in \maybebeta.

@@

\begin{theorem}[Fidelity in $\wellb_{\beta \eta}$]
  % F x = z, F x y = z y
  Given a list of unification problems \foUnifPb such that
  $\wellb_{\beta \eta}(\foUnifPb)$, if the 
  memory map is bijective then
  the introduction 
  of \linketa guarantees \fullref{prop:fidelity}.
  \label{lemma:fidelity-maybeeta}
\end{theorem}

\begin{proof}[Proof sketch]
  We want to prove that \fstep succeeds iff \hstep
  succeeds, where \hstep is made by a unification step $u$ and a link
  progression $p$.
  Without loss of generality we consider a
  \Fo{} unification problem of the form $s_1 \Uo s_2$
  where $s_1 \in \maybeeta$ and is compiled to a variable $X$ with the accessory link
  $\linketaM{\Gamma}{X}{\lambda x.t_1}$. 
  The unification $u$ always succeeds, since $X$ is a fresh variable, we 
  only have to consider the link progression $p$.
  Two cases should be analyzed:
  \begin{itemize}
    \item $s_2 \in \maybeeta$, in this case, the unification problem becomes $X \Ue Y$ 
          with the extra link $\linketaM{}{Y}{\lambda x.t_2}$. After the unification of 
          $X$ and $Y$, \progressetadedup is fired and, once the lambdas are crossed,
          $t_1$ and $t_2$ are unified. 
          This unification succeeds iff $s_1 \Uo s_2$ succeeds by \cref{th:fidelity-beta}.

@@

\section{Making \mapStore a bijection}
\label{sec:invariant1}

We want to allow for partial application of functions.
As a consequence
the same unification variable $X$ can be used multiple times
with different arities.
When it is the case the
memory map \mapStore generated by the compiler is not
a bijection. For example:
% make test ONLY=7007 TEX=tex
\printAlll
  {{{\lambda x.(X\appsep x),f},
    {X,\lambda x.a}}}
  {{{A,f},
    {C,\lambda x.a}}}
  {{{X,C,0},
    {X,B,1}}}
  {{{\eta,,A,\lambda x.B_{x}}}}
in the unification problems \foUnifPb above, $X$ is used with arity $1$ in
$\foUnifPb_1$ and with arity $0$ in $\foUnifPb_2$. 
In order to preserve \fullref{inv:uvaarity} the compiler
did generate two entries in \mapStore for $X$, namely $B$ and $C$. 
However, there is no connection between these two variables
and any incompatible assignements to them would not be detected,
in turn breaking \cref{prop:fidelity}. 
% This incompatibility will only be detected during the
% decompilation phase.
To address this issue, we post-process \mapStore
to ensure the following property:

@@

% Note that shortening the chain of links allows unification to fail if
% $X_1$ and $X_2$ are unified with different terms.

@@

\subsection{\texorpdfstring{Relaxing \fullref{def:progBetaFail}}{Relaxing progress-llam-fail}}
\label{sec:heuristics}
\newcommand{\progBetaNoLLWait}{\emph{progress-beta-\notllambda}}
